Experience Report: Haskell in Computational Biology

### 2012-03-10 mrfy_experience_report.tex
another small verbatim instance
coming dangerously near collision with noah

\newcommand\txprobj[3][]{a#1_{#2_{j-1}#3_j}}
\newcommand\txprobjj[3][]{a#1_{#2_{j-1}#3_j}}

@@

Once a protein alignment is constructed, it~is used to train a
\emph{hidden Markov model}. 
A~hidden Markov model is a probabilistic finite-state machine which 
can assign a probability to any query sequence.
A~protein whose query sequence has a higher probability is more likely to %
be homologous to the proteins in the alignment.
We~write a query sequence as $x_1, \ldots, x_{\scriptscriptstyle N}$,
where each $x_i$~is 
an amino acid.

@@

\caption{A~structural alignment of four proteins}
\label{alignment} 
\end{figure}

A~hidden Markov model carries probabilities on its states and on its
state transitions.
Both the probabilities and the
the states are determined by the alignment:
\begin{itemize}
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}.
The match state contains a table $e_{M_j}(x)$ which gives the
 probability that a homologous protein has amino acid~$x$ in
 column~$j$.
This probability is called an \emph{emission} probability.
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{deletion state}.
The deletion state determines the probability that the query sequence
has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
\item 
For each column~$j$ of the alignment, the hidden Markov model has an
\emph{insertion state}.
The insertion state contains a table $e_{I_j}(x)$ which gives the
probability that a homologous protein has gained amino acid~$x$ by
insertion at column~$j$.
\end{itemize}
The hidden Markov model also has distinguished \emph{begin} and \emph{end} states.

@@

This model, trained from a four-column alignment,
has begin and end states $B$~and~$E$,
as well as four
nodes, each containing 
an insertion state~$I$, 
a match state~$M$, and a
deletion state~$D$.

\caption{A hidden Markov model}

@@

\begin{figure*}
\def\goo{18pt}
\def\gum{14pt}
\[
\begin{array}{l@{}c@{}l}
V_{j}^{M}(i) &{}={}& \log\frac{e_{M_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
\begin{array}{l l}
V_{j-1}^{M}(i - 1) + \log a_{M_{j-1}M_{j}}\\
V_{j-1}^{I}(i - 1) + \log a_{I_{j-1}M_{j}}\\
V_{j-1}^{D}(i - 1) + \log a_{D_{j-1}M_{j}}\\
\end{array} \right.\\[\goo]
V_{j}^{I}(i) &=& \log\frac{e_{I_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
\begin{array}{l l}
V_{j}^{M}(i - 1) + \log a_{M_{j}I_{j}}\\
V_{j}^{I}(i - 1) + \log a_{I_{j}I_{j}}\\
\end{array} \right.\\[\gum]
V_{j}^{D}(i) &=& \max \left\{
\begin{array}{l l}
V_{j-1}^{M}(i) + \log a_{M_{j-1}D_{j}}\\
V_{j-1}^{D}(i) + \log a_{D_{j-1}D_{j}}\\
\end{array} \right.\\
\end{array}
\qquad
\begin{array}{l@{}c@{}l}
V_{j}^{\prime M}(i) &{}={}& e^{\prime}_{M_{j}}(x_{i}) + \min \left\{
\begin{array}{l l}
V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j-1}M_{j}}\\
V_{j-1}^{\prime I}(i - 1) + a^{\prime}_{I_{j-1}M_{j}}\\
V_{j-1}^{\prime D}(i - 1) + a^{\prime}_{D_{j-1}M_{j}}\\
\end{array} \right.\\[\goo]
V_{j}^{\prime I}(i) &=& e^{\prime}_{I_{j}}(x_{i}) + \min \left\{
\begin{array}{l l}
V_{j}^{\prime M}(i - 1) + a^{\prime}_{M_{j}I_{j}}\\
V_{j}^{\prime I}(i - 1) + a^{\prime}_{I_{j}I_{j}}\\
\end{array} \right.\\[\gum]
V_{j}^{\prime D}(i) &=& \min \left\{
\begin{array}{l l}
V_{j-1}^{\prime M}(i) + a^{\prime}_{M_{j-1}D_{j}}\\
V_{j-1}^{\prime D}(i) + a^{\prime}_{D_{j-1}D_{j}}\\
\end{array} \right.\\
\end{array}
\]

\centerline{$a'_{s} = - \log a_{s} \nonumber$,
$e^{\prime}_{s,x} = - \log\frac{e_{s,x}}{q_{x}}$,
and 
$V_j^{\prime M}(i) = - V_j^{M}(i)$}
\caption{Viterbi's equations, in original and negated forms}
\label{viterbi}
\end{figure*}

@@

Given a hidden Markov model, 
an~established software package called HMMER (``hammer'') 
can compute the probability
that a new protein shares structure and
evolutionary history with the proteins used to train the model.
The computation finds the most likely path through the hidden Markov model.
To~make best use of floating-point arithmetic, the software computes
the \emph{logarithm} of the probability of each path, by summing
the logs of the 
probabilities on the states and edges of the path \cite{Viterbi:1967hq}.
(The logarithm function is monotonic, so the path that maximizes the
log of the probability is the most likely path.)
The~specification is shown on the left-hand side of \figref{viterbi}.
The probability $V_j^M(i)$ \emph{[REPRESENTS WHAT EXACTLY?]}.

To~be able to use Haskell, we had to reimplement this standard
algorithm.
Haskell made it possible for us to write code that looks like the
math,
which made the code easy to write and gives us confidence that it is
correct.
We~begin with our representations.

@@

A~transition probability is simply a number.
For accuracy in computation, we add log probabilities instead of
multiplying probabilities.
And to distinguish a transition probability from other numbers, we wrap it in
the value constructor \texttt{TProb}:
\begin{smallverbatim}
type LogProb = Double
data TProb  = TProb { logProbability :: LogProb }
\end{smallverbatim}

We group transition probabilities into 7-tuples.
Each tuple corresponds to a ``node'' in the hidden Markov model
(a~dashed ovel in \figref{plan7}), and to a column of the original
alignment.
As~noted above, 
a match state may transition to any state in the successor node;
an insertion state may transition to itself or to its successor match
state;
and
a~deletion may transition to its successor match
state or deletion state.
\begin{smallverbatim}
data TProbs = 
     TProbs { m_m :: TProb, m_i :: TProb, m_d :: TProb
            , i_m :: TProb, i_i :: TProb
            , d_m :: TProb, d_d :: TProb }
\end{smallverbatim}

The
and a vector of emission probabilities for the match and insertion states.
\begin{smallverbatim}
type EProbs = V.Vector LogProb
data HmmNode = HmmNode { nodeNum :: Int
                       , matEmissions :: EProbs
                       , insEmissions :: EProbs
                       , transitions :: TProbs
                       }
\end{smallverbatim}

The Viterbi algorithm comprises a set of recurrence 
relations (see Equation \ref{viterbi_eqn}) that seek to maximize the 
probability of an observed sequence (in this case, of amino acid ``letters'') 
being emitted by a particular finite-state automaton (the hidden Markov model).

To be efficient, the Viterbi 
algorithm relies on dynamic programming, or memoization, and the resulting 
asymptotic complexity is $O(|M|\times|N|)$, when $M$ is the sequence of 
distinct states (nodes in the automaton's graph) and $N$ is the sequence of 
letters in the protein sequence.
In our particular problem domain, both $M$ and 
$N$ routinely contain several hundred to a few thousand elements.

The textbook Viterbi recurrence relations refer to logs of probabilities, so 
that the terms may be added rather than multiplying raw probabilities.
For 
reasons including floating point underflow, the HMMER software (with which we 
maintain file format compatibility) stores all probabilities in a trained HMM 
file as negative natural logs.
Thus, the Viterbi recurrence relations are 
simplified from the form in \ref{viterbi_eqn} to that in \ref{viterbi_log_eqn}, 
and because they are \textit{negative} logs, the problem transforms from 
maximization to minimization.

We implemented the Viterbi algorithm in two distinct ways.
We first wrote a 
bottom-up dynamic-programming version, which began with the first node of the 
hidden Markov model and first observation and built up the three-dimensional 
array of possible paths through the model.
The array containing the paths 
served as the memoization table.
We next wrote a top-down dynamic-programming 
version, which began with the last node of the model and the last observation, 
and applied structural recursion down to the base cases involving the first 
node or first observation (or both).
The general case of this top-down, structurally recursive implementation closely
resembles the Viterbi equations above.
The match equation is thus:
\begin{smallverbatim}
viterbi s n o
  | s == mat = 
       let trans t prev = 
        (scr + tProb (n-1) t + (eProb n o), mat:p)
        where (scr, p) = viterbi prev (n-1) (o-1)
       in minimum $ [
           trans m_m mat,
           trans i_m ins,
           trans d_m del
           ]
\end{smallverbatim}
% Similarly, the insertion equation is:
% \begin{verbatim}                    
%   -- consume an observation but not a node
%   | s == ins = let trans t prev =
%                    (score + tProb n t + (eProb n o), ins:path)
%                     where (score, path) = viterbi prev n (o - 1)
%                 in minimum [
%                     trans m_i mat,
%                     trans i_i ins
%                     ]
% \end{verbatim}
% Finally, the deletion equation is:
% \begin{verbatim}                    
%   -- consume a node but not an observation
%   | s == del = let trans t prev = 
%                       (score + tProb (n-1) t, del:path)
%                       where (score, path) = viterbi prev (n - 1) o
%                 in minimum [
%                     trans m_d mat,
%                     trans d_d del
%                     ]
% \end{verbatim}

>>>>

\newcommand\txprobj[3][]{a#1_{#2_{j-1}#3_j}}
\newcommand\txprobjj[3][]{a#1_{#2_{j-1}#3_j}}
\newcommand\alignwidth{\ensuremath R} % number of columns in an alignment

@@

Once a protein alignment is constructed, it~is used to train a
\emph{hidden Markov model}. 
A~hidden Markov model is a probabilistic finite-state machine which 
can assign a probability to any query sequence.
A~protein whose query sequence has a higher probability is more likely to %
be homologous to the proteins in the alignment.
We~write a query sequence as $x_1, \ldots, x_{\scriptscriptstyle N}$,
where each $x_i$~is 
an amino acid.
$N$~need not be the same as the number of columns in the alignment,
which we write~\alignwidth.

@@

\caption{A~structural alignment of four proteins $(\alignwidth = 4)$}
\label{alignment} 
\end{figure}

A~hidden Markov model carries probabilities on its states and on its
state transitions.
Both the probabilities and the
the states are determined by the alignment:
\begin{itemize}
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{match state}~$M_j$.
The match state contains a table $e_{M_j}(x)$ which gives the
 probability that a homologous protein has amino acid~$x$ in
 column~$j$.
This probability is called an \emph{emission} probability.
\item
For each column~$j$ of the alignment, the hidden Markov model has a
\emph{deletion state}~$D_j$.
The deletion state determines the probability that the query sequence
has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
\item 
For each column~$j$ of the alignment, the hidden Markov model has an
\emph{insertion state}~$I_j$.
The insertion state contains a table $e_{I_j}(x)$ which gives the
probability that a homologous protein has gained amino acid~$x$ by
insertion at column~$j$.
\end{itemize}
The hidden Markov model also has distinguished \emph{begin} and \emph{end} states.

@@

This model
has begin and end states $B$~and~$E$,
as well as four
nodes, each containing 
an insertion state~$I$, 
a match state~$M$, and a
deletion state~$D$.

\caption{A hidden Markov model $(\alignwidth = 4)$}

@@

\begin{figure*}
\def\goo{18pt}
\def\gum{14pt}
\[
\begin{array}{l@{}c@{}l}
V_{j}^{M}(i) &{}={}& \log\frac{e_{M_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
\begin{array}{l l}
V_{j-1}^{M}(i - 1) + \log a_{M_{j-1}M_{j}}\\
V_{j-1}^{I}(i - 1) + \log a_{I_{j-1}M_{j}}\\
V_{j-1}^{D}(i - 1) + \log a_{D_{j-1}M_{j}}\\
\end{array} \right.\\[\goo]
V_{j}^{I}(i) &=& \log\frac{e_{I_{j}}(x_{i})}{q_{x_{i}}} + \max \left\{
\begin{array}{l l}
V_{j}^{M}(i - 1) + \log a_{M_{j}I_{j}}\\
V_{j}^{I}(i - 1) + \log a_{I_{j}I_{j}}\\
\end{array} \right.\\[\gum]
V_{j}^{D}(i) &=& \max \left\{
\begin{array}{l l}
V_{j-1}^{M}(i) + \log a_{M_{j-1}D_{j}}\\
V_{j-1}^{D}(i) + \log a_{D_{j-1}D_{j}}\\
\end{array} \right.\\
\end{array}
\qquad
\begin{array}{l@{}c@{}l}
V_{j}^{\prime M}(i) &{}={}& e^{\prime}_{M_{j}}(x_{i}) + \min \left\{
\begin{array}{l l}
V_{j-1}^{\prime M}(i - 1) + a^{\prime}_{M_{j-1}M_{j}}\\
V_{j-1}^{\prime I}(i - 1) + a^{\prime}_{I_{j-1}M_{j}}\\
V_{j-1}^{\prime D}(i - 1) + a^{\prime}_{D_{j-1}M_{j}}\\
\end{array} \right.\\[\goo]
V_{j}^{\prime I}(i) &=& e^{\prime}_{I_{j}}(x_{i}) + \min \left\{
\begin{array}{l l}
V_{j}^{\prime M}(i - 1) + a^{\prime}_{M_{j}I_{j}}\\
V_{j}^{\prime I}(i - 1) + a^{\prime}_{I_{j}I_{j}}\\
\end{array} \right.\\[\gum]
V_{j}^{\prime D}(i) &=& \min \left\{
\begin{array}{l l}
V_{j-1}^{\prime M}(i) + a^{\prime}_{M_{j-1}D_{j}}\\
V_{j-1}^{\prime D}(i) + a^{\prime}_{D_{j-1}D_{j}}\\
\end{array} \right.\\[\gum]
\multicolumn3{l}{%
a'_{s} = - \log a_{s} 
\qquad
e^{\prime}_{s,x} = - \log\frac{e_{s,x}}{q_{x}}
\qquad
V_j^{\prime M}(i) = - V_j^{M}(i)
}
\\
\end{array}
\]

\caption{Viterbi's equations, in original and negated forms}
\label{viterbi}
\end{figure*}

@@

Given a hidden Markov model, 
an~established software package called HMMER (``hammer'') 
can compute the probability
that a new protein shares structure and
evolutionary history with the proteins used to train the model.
The computation finds the most likely path through the hidden Markov model.
To~make best use of floating-point arithmetic, the software computes
the \emph{logarithm} of the probability of each path, by summing
the logs of the 
probabilities on the states and edges of the path \cite{Viterbi:1967hq}.
(The logarithm function is monotonic, so the path that maximizes the
log of the probability is the most likely path.)
The~specification is shown on the left-hand side of \figref{viterbi}.
The probability $V_j^M(i)$ represents the probability of the most
likely path of the first~$i$ amino acids in the query sequence,
terminating with placement of amino acid $x_i$ in state~$M_j$.
The probabilities $V_j^D(i)$ and $V_j^D(i)$ are similar.

To~be able to use Haskell, we had to reimplement the standard
algorithm for solving Viterbi's equations.
Haskell made it possible for us to write code that looks like the
math,
which made the code easy to write and gives us confidence that it is
correct.
We~begin with our representations.

@@

The log of a probability is simply a number.
To distinguish a transition probability from other numbers, we wrap it in
the value constructor \texttt{TProb}:
\begin{smallverbatim}
type LogProb = Double
data TProb   = TProb { logProbability :: LogProb }
\end{smallverbatim}
We group transition probabilities into 7-tuples.
Each tuple corresponds to a ``node'' in the hidden Markov model
(a~dashed ovel in \figref{plan7}), and to a column of the original
alignment.
As~noted above, 
a match state may transition to any state in the successor node;
an insertion state may transition to itself or to its successor match
state;
and
a~deletion may transition to its successor match
state or deletion state.
\begin{smallverbatim}
data TProbs = 
     TProbs { m_m :: TProb, m_i :: TProb, m_d :: TProb
            , i_m :: TProb, i_i :: TProb
            , d_m :: TProb, d_d :: TProb }
\end{smallverbatim}

The representation of a node includes the probabilities of transitions
into the three states of that node, plus the tables
of emission probabilities for the match and insertion states.
\begin{smallverbatim}
type EProbs  = Vector LogProb
data HmmNode = HmmNode { transitions  :: TProbs
                       , matEmissions :: EProbs
                       , insEmissions :: EProbs
                       }
\end{smallverbatim}

In~Viterbi's equations,
the probability in each state is a simple function of the probability
in its predecessor states, 
and all probabilities can be computed by a classic dynamic-programming
algorithm
which starts at the begin state,
computes probabilites in nodes $1$~through~$\alignwidth$ in
succession, and terminates at the end state.
We~implemented this algorithm, storing the probabilities in an array.
\remark{Why is the number of dimensions 3 and not 2 (indices $i$, $j$,
and what??) and why do we care?}
The cost is
$O(|N|\times|\alignwidth|)$;
in our work, $\alignwidth$~and~$N$ range from several hundred to a few
thousand.

%%  For 
%%  reasons including floating point underflow, the HMMER software (with which we 
%%  maintain file format compatibility) stores all probabilities in a trained HMM 
%%  file as negative natural logs.
%%  Thus, the Viterbi recurrence relations are 
%%  simplified from the form in \ref{viterbi_eqn} to that in \ref{viterbi_log_eqn}, 
%%  and because they are \textit{negative} logs, the problem transforms from 
%%  maximization to minimization.

\emph{[WHAT WAS UNSATISFACTORY ABOUT THE FIRST IMPLEMENTATION]}

We next wrote a top-down dynamic-programming 
version, which began with the last node of the model and the last observation, 
and applied structural recursion down to the base cases involving the first 
node or first observation (or both).
The general case of this top-down, structurally recursive implementation closely
resembles the Viterbi equations above.
The match equation is thus:
\begin{smallverbatim}
viterbi s n o
  | s == mat = 
       let trans t prev = 
        (scr + tProb (n-1) t + (eProb n o), mat:p)
        where (scr, p) = viterbi prev (n-1) (o-1)
       in minimum $ [
           trans m_m mat,
           trans i_m ins,
           trans d_m del
           ]
\end{smallverbatim}
% Similarly, the insertion equation is:
% \begin{verbatim}                    
%   -- consume an observation but not a node
%   | s == ins = let trans t prev =
%                    (score + tProb n t + (eProb n o), ins:path)
%                     where (score, path) = viterbi prev n (o - 1)
%                 in minimum [
%                     trans m_i mat,
%                     trans i_i ins
%                     ]
% \end{verbatim}
% Finally, the deletion equation is:
% \begin{verbatim}                    
%   -- consume a node but not an observation
%   | s == del = let trans t prev = 
%                       (score + tProb (n-1) t, del:path)
%                       where (score, path) = viterbi prev (n - 1) o
%                 in minimum [
%                     trans m_d mat,
%                     trans d_d del
%                     ]
% \end{verbatim}
